A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-10-09 paper/paper.experiments.tex
Update paper.experiments.tex
SM reviewed experiments section

% JK: I think give an overview here of the experiments here
%     and what you're going to show in this section (and why)
% check verb tense - use past tense I think given flow
Our goal was to determine if the schedules provided by solving the ILP 
could successfully
enforce all hard constraints; improve satisfiability of soft constraints compared with ??; %SM: found "improve satisfiability" a bit hard to follow. revise this phrase for clarity
and assess if our ILP formulation can be used for a wide range
of configurations.  %list objectives with numbers (i, ii, iii) or separated by semi-colons - helps the reader re: flow
First, we compared the schedules created by solving the ILP formulation
given in Section~\ref{sec:methods} to schedules that were manually generated,
with respect to adherence to the hard and soft constraints outlined in Section~\ref{sec:problem}.
We then examined the efficiency of the ILP approach in generating schedules
by its runtime on a variety of instances that may be found in the real-world.

@@

\begin{figure}[h]
	\centering
	\def\svgwidth{\columnwidth}
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
	\input{fig/runtime-requests.pdf_tex} % JK: can you export this to PDF or SVG to avoid pixelation?
  \label{fig:runtime-requests}
  % JK: The trend here is a bit confusing, since I think you're trying to say
  %     that the runtime is not affected by the number of constraints,
  %     but there is noise in these results.
  %     I would re-run this 10-100 times and average the results so that (I expect)
  %     the mean trend is actually flat. You could (should) then also present the runtimes
  %     as box-plots to show that the trend is less than the variability.
\end{figure}

@@

>>>>

% JK: I think give an overview here of the experiments here
%     and what you're going to show in this section (and why)
% check verb tense - use past tense I think given flow
Our goal was to determine if the schedules provided by solving the ILP 
could successfully
(i) enforce all hard constraints; 
(ii) improve fulfillment of soft constraints compared to the manual approach; %SM: found "improve satisfiability" a bit hard to follow. revise this phrase for clarity
and (iii) assess whether our ILP formulation can be used for a wide range
of configurations.  %list objectives with numbers (i, ii, iii) or separated by semi-colons - helps the reader re: flow
First, we compared the schedules created by solving the ILP formulation
given in Section~\ref{sec:methods} to schedules that were manually generated,
with respect to adherence to the hard and soft constraints outlined in Section~\ref{sec:problem}.
We then examined the efficiency of the ILP approach in generating schedules
by its runtime on a variety of instances that may be found in the real-world.

@@

\begin{figure}[h]
	\centering
	\def\svgwidth{\columnwidth}
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
	\input{fig/runtime-requests.pdf_tex} % JK: can you export this to PDF or SVG to avoid pixelation?
  \label{fig:runtime-requests}
  % JK: The trend here is a bit confusing, since I think you're trying to say
  %     that the runtime is not affected by the number of constraints,
  %     but there is noise in these results.
  %     I would re-run this 10-100 times and average the results so that (I expect)
  %     the mean trend is actually flat. You could (should) then also present the runtimes
  %     as box-plots to show that the trend is less than the variability.  %SM: great suggestion! Did you do this (re-run 10-100 times?) - was not clear from description of experiments above
\end{figure}

@@

%SM: Table 6. change column heading "number of services" to "number of services within 1 division"
%SM: Table 6. without NCB is a double-negative (without no consecutive blocks). Can this be revised?

### 2019-10-09 paper/paper.methods.tex

\subsection{Constraints}\label{sec:meth-constraints}
We now formalize the hard constraints in Table~\ref{tbl:constraint-summary}
using the variables defined above.

% JK: not sure how I feel about these letters as equation refs
% SM: agree. on reading in full, it felt odd enough for me to suggest using number for equations...
% e.g. The WC (weekend coverage) constraint is given by .... [equation]
\begin{align}
&\sum_{c=1}^{C} X_{c, b, s} = 1 &&\forall b\in \mathcal{B}, s \in \mathcal{S}
\tag{BC} \label{eqn:constr-block-cov} \\
&\sum_{c=1}^{C} Y_{c, w} = 1 &&\forall w\in \mathcal{W} \tag{WC}
\label{eqn:constr-weekend-cov} \\
&m_{c, s} \leq \sum_{b=1}^{B} X_{c, b, s} \leq M_{c, s} &&\forall\
c\in\mathcal{C}, s\in\mathcal{S} \tag{MM} \label{eqn:constr-min-max} \\
&X_{c, b, s} + X_{c, b + 1, s} \leq 1 &&\forall c\in\mathcal{C}, b \leq B - 1,
s\in\mathcal{S} \tag{NCB} \label{eqn:constr-no-consec-blocks} \\
&Y_{c, w} + Y_{c, w + 1} \leq 1 &&\forall c\in\mathcal{C}, w \leq W - 1
\tag{NCW} \label{eqn:constr-no-consec-weekends} \\
&\floor*{\frac{W}{C}} \leq \sum_{w=1}^W Y_{c, w} \leq \ceil*{\frac{W}{C}}
&&\forall c\in\mathcal{C} \tag{EW} \label{eqn:constr-equal-weekends} \\
&\floor*{\frac{\abs{\mathcal{L}}}{C}} \leq \sum_{w\in\mathcal{L}} Y_{c, w} \leq
\ceil*{\frac{\abs{\mathcal{L}}}{C}} &&\forall c\in\mathcal{C} \tag{EH}
\label{eqn:constr-equal-holidays}
\end{align}

\subsection{Objectives}\label{sec:meth-objectives}
As described in Section~\ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section~\ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions of $X$ and $Y$:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer  %SM: agree!
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
{(-1)}^{\ind(b\,\in\,\mathcal{U}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
{(-1)}^{\ind(w\,\in\,\mathcal{V}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments.

The Block-Weekend Adjacency is optimized by considering the product $X_{c, b,
	s}\cdot Y_{c, w}$ for values of $w$ ``adjacent'' to the value of $b$. This
leads
to the maximization objective %avoid editorial adjectives
\begin{align}
&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} X_{c, b, s}\cdot Y_{c,
	w=\varphi(b)} \tag{BWA} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ is a one-to-one mapping of a block to an adjacent weekend, by
some appropriate definition of adjacency. For instance, clinicians might want to
be assigned during a weekend that falls within an assigned block. In this case,
we will have $\varphi(b) = 2b-1$.

>>>>

\subsection{Constraints}\label{sec:meth-constraints}
We now formalize the hard constraints in Table~\ref{tbl:constraint-summary}
using the variables defined above. 
The BC (block coverage) and WC (weekend coverage) constraints, 
are given by Eqns. (\ref{eqn:constr-block-cov}) and (\ref{eqn:constr-weekend-cov}), respectively. 
\begin{align}
&\sum_{c=1}^{C} X_{c, b, s} = 1 &&\forall b\in \mathcal{B}, s \in \mathcal{S} \label{eqn:constr-block-cov} \\
&\sum_{c=1}^{C} Y_{c, w} = 1 &&\forall w\in \mathcal{W} \label{eqn:constr-weekend-cov}
\end{align}
The MM (min/max) constraint is given by:
\begin{align}
&m_{c, s} \leq \sum_{b=1}^{B} X_{c, b, s} \leq M_{c, s} &&\forall\
c\in\mathcal{C}, s\in\mathcal{S} \label{eqn:constr-min-max}
\end{align}
The NCB (no consecutive blocks) and NCW (no consecutive weekends) constraints are 
given by Eqns. (\ref{eqn:constr-no-consec-blocks}) and (\ref{eqn:constr-no-consec-weekends}), respectively.
\begin{align}
&X_{c, b, s} + X_{c, b + 1, s} \leq 1 &&\forall c\in\mathcal{C}, b \leq B - 1,
s\in\mathcal{S} \label{eqn:constr-no-consec-blocks} \\
&Y_{c, w} + Y_{c, w + 1} \leq 1 &&\forall c\in\mathcal{C}, w \leq W - 1 \label{eqn:constr-no-consec-weekends}
\end{align}
The EW (equal weekend) and EH (equal holidays) constraints are given by Eqns. (\ref{eqn:constr-equal-weekends})
and (\ref{eqn:constr-equal-holidays}), respectively.
\begin{align}
&\floor*{\frac{W}{C}} \leq \sum_{w=1}^W Y_{c, w} \leq \ceil*{\frac{W}{C}}
&&\forall c\in\mathcal{C} \label{eqn:constr-equal-weekends} \\
&\floor*{\frac{\abs{\mathcal{L}}}{C}} \leq \sum_{w\in\mathcal{L}} Y_{c, w} \leq
\ceil*{\frac{\abs{\mathcal{L}}}{C}} &&\forall c\in\mathcal{C}
\label{eqn:constr-equal-holidays}
\end{align}
% JK: not sure how I feel about these letters as equation refs
% SM: agree. on reading in full, it felt odd enough for me to suggest using number for equations...
% e.g. The WC (weekend coverage) constraint is given by .... [equation]

\subsection{Objectives}\label{sec:meth-objectives}
As described in Section~\ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section~\ref{sec:meth-variables}. 
Objectives BR and WR are given in Eqns. (\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests})
as linear functions of $X$ and $Y$:
%minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer  %SM: agree!
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
{(-1)}^{\ind(b\,\in\,\mathcal{U}_c)}\cdot X_{c, b, s}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
{(-1)}^{\ind(w\,\in\,\mathcal{V}_c)}\cdot Y_{c, w}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments.

The BWA objective is optimized by considering the product $X_{c, b, s}\cdot Y_{c, w}$ 
for values of $w$ ``adjacent'' to the value of $b$. This
leads to the maximization objective:%avoid editorial adjectives
\begin{align}
&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} X_{c, b, s}\cdot Y_{c,
	w=\varphi(b)} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ is a one-to-one mapping of a block to an adjacent weekend, by
some appropriate definition of adjacency. For instance, clinicians might want to
be assigned during a weekend that falls within an assigned block. In this case,
we will have $\varphi(b) = 2b-1$.

### 2019-10-09 paper/tbl/runtime_services_clinicians.tex

\begin{table}[htbp]
	\centering
 	\caption{Comparison of program runtime (in seconds) for different numbers of services and total clinicians in the division.}%
  \label{tbl:runtime-services-clinicians-comparison}%
	\begin{tabular}{|c|c||c|c||c|c|}
		\toprule
		                                      &  \multicolumn{5}{c|}{Number of Services}  \\ \midrule
		\makecell[l]{Number of \\ Clinicians} &  1   &    2    & 2 (without NCB) &  3   & 3 (without NCB) \\ \midrule
		                 10                   & 0.16 &  0.74   &  0.16   & 1.40 &  0.23   \\ \hline
		                 20                   & 0.25 & 7468.86 &  0.32   &  --  &  0.43   \\ \hline
		                 30                   & 0.42 &    --   &  0.49   &  --  &  0.66   \\ \hline
		                 50                   & 0.62 &    --   &  0.82   &  --  &  1.14   \\ \bottomrule
	\end{tabular}\\[1em]
  \footnotesize\raggedright
  Notes:
  ``--'' indicates that no solution was found within 24 hours;
  ``without NCB'' indicates that the No Consecutive Blocks (NCB) constraint was removed.
\end{table}

>>>>

\begin{table}[htbp]
	\centering
 	\caption{Comparison of program runtime (in seconds) for different numbers of services and total clinicians in the division.}%
  \label{tbl:runtime-services-clinicians-comparison}%
	\begin{tabular}{|c|c||c|c||c|c|}
		\toprule
		                                      &  \multicolumn{5}{c|}{Number of Services}  \\ \midrule
		\makecell[l]{Number of \\ Clinicians} &  1   &    2    & 2 (*) &  3   & 3 (*) \\ \midrule
		                 10                   & 0.16 &  0.74   &  0.16   & 1.40 &  0.23   \\ \hline
		                 20                   & 0.25 & 7468.86 &  0.32   &  --  &  0.43   \\ \hline
		                 30                   & 0.42 &    --   &  0.49   &  --  &  0.66   \\ \hline
		                 50                   & 0.62 &    --   &  0.82   &  --  &  1.14   \\ \bottomrule
	\end{tabular}\\[1em]
  \footnotesize\raggedright
  Notes:
  ``--'' indicates that no solution was found within 24 hours;
  ``(*)'' indicates that the No Consecutive Blocks (NCB) constraint was removed.
\end{table}
